---
title: "Generative model simulations"
output: html_document
---
# Loading some packages and additional scripts
```{r}
source("function.R")
library(lfmm)
library(cate)
library(glmnet)
library(sva)
library(tidyverse)
library(reshape2)
```

#  Running generative simulations

This section reproduces the experiments described in section "Generative simulation experiments" of the manuscript. Generative simulations are based on the generative model of LFMM, explained in the main text equation (1).

## Loading the data and setting simulation parameters

Here we define the parameters used in the simulations. 

```{r}
# Number of individuals
n <- 400
# n <- 1000

# Number of genetic markers
p <- 500 
# p <- 20000 # warning, take a longer time to run

# Number of latent factors
k <- 6

# Number of causal genetic markers
nc <- 80 

# Proportion of causal markers
pc <- nc/p

# Evaluated simulation parameters
# Proportion of exposure variance explained by latent structure
# also defined as 'intensity of confounding'
prv <- rep(c(0.1, 0.5), 3)
# prv <- rep(seq(0.1, 0.7, 0.2), 200)

# mean of effect sizes for causal markers
eff <- c(0.75, 1.5)

para <- expand.grid(prv, eff)
colnames(para) <- c("prop_var", "eff")

# Number of simulations
nb <- nrow(para)

# Performance matrices
# F1-score
f1 <- matrix(NA, nrow = nb, ncol = 5)

# Root Mean Square Error computed on all markers
rmse <- matrix(NA, nrow = nb, ncol = 6)

# Root Mean Square Error computed on causal markers
rmse.c <- matrix(NA, nrow = nb, ncol = 5)

# Precision (1 - False Discovery Rate)
preci <- matrix(NA, nrow = nb, ncol = 5)

# Recall (Power)
recal <- matrix(NA, nrow = nb, ncol = 5)
```

## Creating simulated data sets and evaluating methods 

For each simulated dataset, we performed association analyses using six methods: sparse lfmm, bslmm, lasso regression, ridge lfmm, cate and sva. All methods except bslmm are implemented in R, and bslmm (see **Estimation algorithms** in the text). 

For each method, we computed the following performance measures: F1-score, precision and recall.


```{r}
for (j in 1:nrow(para)) {
  
# Create simulated data 
    kr <- 4
    simu <- simulator_ewas(n = n, p = p, K = k + kr, prop.causal = pc, 
                           prop.variance = para$prop_var[j], mean.A = para$eff[j], 
                           sd.V = 1, sd.U = c(runif(k, 2, 6), runif(kr, 0, 1)))

    # get the subset of causal markers
    causal <- simu$causal

# For each method, record association values (effect sizes, B)
    
    # sparse lfmm
    slf <- lfmm::lfmm_lasso(Y = simu$Y, X = simu$X, K = k, nozero.prop = pc)$B

    # lasso from glmnet package
    cvfit <- glmnet::cv.glmnet(x = simu$Y, y = simu$X)
    a <- coef(cvfit, s = cvfit$lambda.1se) 
    b <- data.frame(a@i, a@x)[-1,] # remove intercept
    las <- rep(0, p)
    las[b[,1]] <- b[,2]

    # lfmm ridge
    rlf <- lfmm::lfmm_ridge(Y = simu$Y, X = simu$X, K = k)$B

    # cate 
    cat <- cate::cate.fit(X.primary = simu$X, Y = simu$Y, r = k)$beta

    # sva
    mod <- model.matrix(~ simu$X)
    mod0 <- model.matrix(~ 1, data = data.frame(simu$X))
    svobj <- sva::sva(t(simu$Y), mod, mod0, n.sv = k)
    modSv <- cbind(mod, svobj$sv)
    sva <- apply(simu$Y, 2, function(x) summary(lm(x ~ modSv[,-1]))$coefficients[2, 1])
    
 
# Compute F1 score, precision and recall
    a.slf <- F1(slf, causal, nb.hit = 100)
    a.las <- F1(las, causal, nb.hit = 100)
    a.rlf <- F1(rlf, causal, nb.hit = 100)
    a.cat <- F1(cat, causal, nb.hit = 100)
    a.sva <- F1(sva, causal, nb.hit = 100)
    
# Compute RMSE 
    r.slf <- RMSE(slf, simu$A)
    r.las <- RMSE(las, simu$A)
    r.rlf <- RMSE(rlf, simu$A)
    r.cat <- RMSE(cat, simu$A)
    r.sva <- RMSE(sva, simu$A)
    r.zer <- RMSE(rep(0, p), simu$A)
    
# Compute RMSE on causal markers
    rc.slf <- RMSE(slf[causal], simu$A[causal])
    rc.las <- RMSE(las[causal], simu$A[causal])
    rc.rlf <- RMSE(rlf[causal], simu$A[causal])
    rc.cat <- RMSE(cat[causal], simu$A[causal])
    rc.sva <- RMSE(sva[causal], simu$A[causal])

# Record results
    # F1-score
    f1[j,] <- c(a.slf$f1, a.las$f1, a.rlf$f1, a.cat$f1, a.sva$f1)
    
    # Precision
    preci[j,] <- c(a.slf$precision, a.las$precision,a.rlf$precision, a.cat$precision, a.sva$precision)
    
    # Recall
    recal[j,] <- c(a.slf$recall, a.las$recall, a.rlf$recall, a.cat$recall, a.sva$recall)
    
    # Root Mean Square Error computed on all markers
    rmse[j,] <- c(r.slf, r.las, r.rlf, r.cat, r.sva, r.zer)
    
    # Root Mean Square Error computed on causal markers
    rmse.c[j,] <- c(rc.slf, rc.las, rc.rlf, rc.cat, rc.sva)
    
    print(paste0("=============="))
    print(paste0("============== ", j / nrow(rmse) * 100, " %"))
    print(paste0("=============="))

}

# Format results
colnames(rmse) <- c("sparse_lfmm", "lasso", "ridge_lfmm", "cate", "sva", "zero")
colnames(rmse.c) <- c("sparse_lfmm", "lasso", "ridge_lfmm", "cate", "sva")
colnames(f1) <- c("sparse_lfmm", "lasso", "ridge_lfmm", "cate", "sva")
colnames(preci) <- c("sparse_lfmm", "lasso", "ridge_lfmm", "cate", "sva")
colnames(recal) <- c("sparse_lfmm", "lasso", "ridge_lfmm", "cate", "sva")

rmse <- data.frame(para, rmse)
rmse.c <- data.frame(para, rmse.c)
f1 <- data.frame(para, f1)
preci <- data.frame(para, preci)
recal <- data.frame(para, recal)

# res <- list(f1 = f1, rmse = rmse, preci = preci, recal = recal, rmse.c = rmse.c)
```

# Script for Figure 1

Figure 1 describes the Root Mean Square Error (RMSE) as a function of the effect size of causal markers and confounding intensity. Two sparse methods (sparse LFMM, LASSO) and three non-sparse methods (ridge LFMM, CATE and SVA) were compared. The “Zero” value corresponds to an RMSE obtained with all effect sizes set to zero (null-model error). 

The following combinations of simulation parameters were explored: (A) Lower effect sizes and confounding intensities, (B) Lower effect sizes and higher confounding intensities, (C) Higher effect sizes and lower confounding intensities, (D) Higher effect sizes and confounding intensities.

```{r}
rmse <- melt(rmse, id.vars = 1:2)
# Confounding intensity transformed as a factor
# 0.1 --> Low Proportion of variance
# 0.5 --> High Proportion of variance
rmse$prop_var <- as.factor(rmse$prop_var)
levels(rmse$prop_var) <- c("LOW \n CONFOUNDING", "HIGH \n CONFOUNDING")

# Effect size transformed as a factor
# 0.75 --> Low EFFECT SIZE
# 1.5 --> High EFFECT SIZE
rmse$eff <- as.factor(rmse$eff)
levels(rmse$eff) <- c("LOW \n EFFECT \n SIZE", "HIGH \n EFFECT \n SIZE")
levels(rmse$variable) <- c("Sparse \n LFMM", "LASSO", "Ridge \n LFMM", "CATE", "SVA", "Zero")

# Summarize the results
dat_mean <- rmse %>%
  group_by(variable, eff, prop_var) %>%
  summarize(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE))

# plot the results
ggplot(dat_mean, aes(variable, mean, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.5, position = position_dodge(0.9)) +
  facet_grid(eff ~ prop_var, scales = "free_x") +
  xlab("INTENSITY OF CONFOUNDING") +
  ylab("RMSE") +
  labs(fill = "METHODS") +
  theme_bw()+
  theme(strip.text.x = element_text(size = 10, face = "bold", angle = 0),
        strip.text.y = element_text(size = 10, face = "bold", angle = 0),
        strip.background = element_rect(fill = "white"),
        axis.title.y = element_text(size = 15, face = "bold"),
        axis.text.y = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 14),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none")+
  scale_fill_manual(values = c("orange", "purple", "blue", "green", "yellow", "black"))

```


# Script for Figure 2

Figure 2 describes the F-score and the precision as a function of effect size of causal markers and confounding intensity. Two sparse methods (sparse LFMM, LASSO) and three non-sparse methods (ridge LFMM, CATE and SVA) were compared. 

The following combinations of simulation parameters were explored: (A) Lower effect sizes and confounding intensities, (B) Lower effect sizes and higher confounding intensities, (C) Higher effect sizes and lower confounding intensities, (D) Higher effect sizes and confounding intensities.

```{r}
dat <- rbind(cbind(f1, stat = "F1 SCORE"),
             cbind(recal, stat = "RECALL"),
             cbind(preci, stat = "PRECISION"))

dat <- melt(dat, id.vars = c(1:2, 8))

# confounding intensity transformed as factor
# 0.1 --> Low Proportion of variance
# 0.5 --> High Proportion of variance
dat$prop_var <- as.factor(dat$prop_var)
levels(dat$prop_var) <- c("LOW \n CONFOUNDING", "HIGH \n CONFOUNDING")

# Effect size transformed asfactor
# 0.75 --> Low EFFECT SIZE
# 1.5 --> High EFFECT SIZE
dat$eff <- as.factor(dat$eff)
levels(dat$eff) <- c("LOW \n EFFECT \n SIZE", "HIGH \n EFFECT \n SIZE")
levels(dat$variable) <- c("Sparse \n LFMM", "LASSO", "Ridge \n LFMM", "CATE", "SVA")

# Summarize and plot the results
dat_mean <- dat %>%
  group_by(variable, eff, prop_var, stat) %>%
  summarize(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE))

filter(dat_mean, stat != "RECALL") %>%
  ggplot(., aes(variable, mean, fill = stat)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.5, position = position_dodge(0.9)) +
  facet_grid(eff ~ prop_var, scales = "free_x") +
  xlab("INTENSITY OF CONFOUNDING") +
  ylab("STATISTICS") +
  labs(fill = NULL) +
  theme_bw()+
  theme(strip.text.x = element_text(size = 10, face = "bold", angle = 0),
        strip.text.y = element_text(size = 10, face = "bold", angle = 0),
        strip.background = element_rect(fill = "white"),
        axis.title.y = element_text(size = 15, face = "bold"),
        axis.text.y = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 14, angle = 0),
        axis.title.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "bottom") +
  #scale_fill_manual(values = c("orange", "yellow", "blue", "green", "purple", "black")) +
  scale_fill_manual(values = c("black", "grey")) +
  scale_y_continuous(limits = c(0, 1))
```

